---
title: "Code for longitudinal SBP and SHT analysis in cats"
author: "C Pye"
date: "2026-06-18"
output: html_document
---

#This R markdown file contains code used to examine changes in SBP and hypertension risk in 209 cats enrolled on CatPAWS. final edit to accompany manuscript

#1. Load libraries
```{r libraries}
library(lme4)
library("ggplot2")
library(splines)
library(dplyr)
library(sjPlot)
library(tidyverse)
library(mclogit)
library(performance)
library(gtsummary)
library(ggfortify)
library(ggridges)
library(patchwork)
library(see)
library(geepack)
library("survival")
library("ggsurvfit")
library("gee")
library("car")
library(boot)
library(rms)

```

#2. load data

```{r data}

Metadata_209_cats<-read.csv("Metadata_hypertension.csv")

```

#3. order factors in df
```{r factors, include = FALSE}

#order OE factors

Metadata_209_cats$oe.abnormalities_Yes_no_incomplete<-as.factor(Metadata_209_cats$oe.abnormalities_Yes_no_incomplete)

levels(Metadata_209_cats$oe.abnormalities_Yes_no_incomplete)
Metadata_209_cats$oe.abnormalities_Yes_no_incomplete<-factor(Metadata_209_cats$oe.abnormalities_Yes_no_incomplete, levels= c("No", "Incomplete_OE", "Yes"))
levels(Metadata_209_cats$oe.abnormalities_Yes_no_incomplete)

```

#4. create dataframe removing SBP NA

```{r create SBP dataset}
Metadata_SBP<-Metadata_209_cats%>%
  filter(!is.na(Metadata_209_cats$SBP))

##remove rows from cats recieving antihypertensives

#8 rows with cats receiving amlodipine
amlodipine_cases<-Metadata_SBP[c(226, 227, 228, 229, 293, 601, 623, 842),]
View(amlodipine_cases)
#8 visits from 5 cats in total. 

#remove these from SBP data

Metadata_SBP2<-Metadata_SBP[c(-226, -227, -228, -229, -293, -601, -623, -842),]

#write csv

#write.csv(Metadata_SBP2, "Metadata_SBP2.csv")
```

#5. summarise data

##a. no.of visits and no.of cats
```{r summary cats and visits, echo=FALSE}

dim(Metadata_SBP2)

Metadata_SBP2%>%
  group_by(Cat_ID)%>%
  summarise(n_distinct(Cat_ID))

```

##b. summarise age data
```{r summary age}

# Function to summarise age stats
summarise_age <- function(data, group_var) {
  data %>%
    group_by({{ group_var }}) %>%
    summarise(
      N = n_distinct(Cat_ID),  # use your unique cat ID column
      Median_Age = round(median(Age_years, na.rm = TRUE), 1),
      IQR_Age = paste0(round(quantile(Age_years, 0.25, na.rm = TRUE), 1), "–", 
                       round(quantile(Age_years, 0.75, na.rm = TRUE), 1)),
      Age_Range = paste0(round(min(Age_years, na.rm = TRUE), 1), "–", 
                         round(max(Age_years, na.rm = TRUE), 1)),
      .groups = "drop"
    )
}

# Summary tables
summarise_age(Metadata_SBP2)

age_by_sex <- summarise_age(Metadata_SBP2, Sex)

age_by_health <- summarise_age(Metadata_SBP2, Overall_health_status)

age_by_breed <- summarise_age(Metadata_SBP2, Breed)

# Combine for full table (optional)
full_age_summary <- bind_rows(
  age_by_sex %>% mutate(Group = paste0("Sex: ", Sex)) %>% select(Group, everything(), -Sex),
  age_by_health %>% mutate(Group = paste0("Health: ", Overall_health_status)) %>% select(Group, everything(), -Overall_health_status),
  age_by_breed %>% mutate(Group = paste0("Breed: ", Breed)) %>% select(Group, everything(), -Breed)
)

# View full summary
print(full_age_summary)

# Summary tables for SBP

age_by_sex <- summarise_age(Metadata_SBP2, Sex)

age_by_health <- summarise_age(Metadata_SBP2, Overall_health_status)

age_by_breed <- summarise_age(Metadata_SBP2, Breed)

# Combine for full table (optional)
full_age_summary <- bind_rows(
  age_by_sex %>% mutate(Group = paste0("Sex: ", Sex)) %>% select(Group, everything(), -Sex),
  age_by_health %>% mutate(Group = paste0("Health: ", Overall_health_status)) %>% select(Group, everything(), -Overall_health_status),
  age_by_breed %>% mutate(Group = paste0("Breed: ", Breed)) %>% select(Group, everything(), -Breed)
)

# View full summary
print(full_age_summary)


# Combine for full table (optional)
full_age_summary <- bind_rows(
  age_by_sex %>% mutate(Group = paste0("Sex: ", Sex)) %>% select(Group, everything(), -Sex),
  age_by_health %>% mutate(Group = paste0("Health: ", Overall_health_status)) %>% select(Group, everything(), -Overall_health_status),
  age_by_breed %>% mutate(Group = paste0("Breed: ", Breed)) %>% select(Group, everything(), -Breed)
)

# View full summary
print(full_age_summary)
```


##c. ridge plots
```{r ridge plots, fig.cap = "Data distribution by age, sex, breed and health status"}

# Plot A: Age by Sex
pA <- ggplot(Metadata_SBP2, aes(x = Age_years, y = Sex, fill = Sex)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  theme_minimal() +
  labs(title = "A. Age distribution by Sex", x = "Age (years)", y = "") +
  theme(legend.position = "none")

# Plot B: Age by Breed
pB <- ggplot(Metadata_SBP2, aes(x = Age_years, y = Breed, fill = Breed)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  scale_fill_manual(values=c("Pedigree" = "yellow", "Non-pedigree" = "green"))+
  theme_minimal() +
  labs(title = "B. Age distribution by Breed", x = "Age (years)", y = "") +
  theme(legend.position = "none")

# Plot C: Age by Health Status
#pC <- ggplot(Metadata_SBP2, aes(x = Age_years, y = #Overall_health_status, fill = Overall_health_status)) +
#  geom_density_ridges(alpha = 0.7, scale = 1.2) +
#  theme_minimal() +
#  labs(title = "C. Age distribution by Overall Health Status", #x = "Age (years)", y = "") +
#  theme(legend.position = "none")

# Plot C: Sex by Breed
pC <- ggplot(Metadata_SBP2, aes(x = Sex, y = Breed, fill = Breed)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  scale_fill_manual(values=c("Pedigree" = "yellow", "Non-pedigree" = "green"))+
  theme_minimal() +
  labs(title = "C. Sex distribution by Breed", x = "Sex", y = "") +
  theme(legend.position = "none")

# Plot D: Age by Sex and Breed
pD <- ggplot(Metadata_SBP2, aes(x = Age_years, fill = Sex)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Breed) +
  theme_minimal() +
  labs(title = "D. Age distribution by Sex and Breed", x = "Age (years)", y = "Density")

# Plot F: Age by Sex and Health Status
pF <- ggplot(Metadata_SBP2, aes(x = Age_years, fill = Sex)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Overall_health_status) +
  theme_minimal() +
  labs(title = "F. Age distribution by Sex and Health Status", x = "Age (years)", y = "Density")

# Plot G: Age by Breed and Health Status
pG<-ggplot(Metadata_SBP2, aes(x = Age_years, fill = Breed)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Overall_health_status) + 
  theme_minimal() +
  labs(title = "G. Age Distribution by Breed and Health Status",
       x = "Age (years)", y = "Density")

# Combine using patchwork
(pA + pB) / (pC+ pD) +
  plot_annotation(title = "Distribution of Age, Sex and Breed",
                  theme = theme(plot.title = element_text(hjust = 0.5, size = 16)))

#ggsave("Supplementary_Figure_DistributionPlots2.png", #width = 12, height = 12, dpi = 300)


```


#6. Analysis of SBP

##a. Summarise data

```{r summary SBP}

# Count the number of visits with sbpdata
num_visits_with_SBP <- Metadata_SBP2 %>%
  filter(!is.na(SBP)) %>%
  summarise(num_visits = n())

# Count the number of cats with sbp data
num_cats_with_SBP <- Metadata_SBP2 %>%
  filter(!is.na(SBP)) %>%
  summarise(num_cats = n_distinct(Cat_ID))

# Print the results
print(num_visits_with_SBP)
print(num_cats_with_SBP)

# Count the number of visits per cat with sbp data
visits_per_cat <- Metadata_SBP2 %>%
  filter(!is.na(SBP)) %>%
  group_by(Cat_ID) %>%
  summarise(num_visits = n())

# Calculate median, IQR, and range
median_visits <- median(visits_per_cat$num_visits)
iqr_visits <- IQR(visits_per_cat$num_visits)
q1 <- quantile(visits_per_cat$num_visits, 0.25)  # First quartile (Q1)
q3 <- quantile(visits_per_cat$num_visits, 0.75)  # Third quartile (Q3)
range_visits <- range(visits_per_cat$num_visits)  # Returns min and max

# Print the results
print(median_visits)
print(iqr_visits)
print(q1)
print(q3)
print(range_visits)

#median follow up time
Metadata_SBP2$date <- as.Date(
  Metadata_SBP2$date,
  format = "%d/%m/%Y"
)
cat_times <- Metadata_SBP2 %>%
  group_by(Cat_ID) %>%
  summarise(
    first_date = min(date, na.rm = TRUE),
    last_date  = max(date, na.rm = TRUE),
    diff_years = as.numeric(difftime(last_date, first_date, units = "days")) / 365.25,
    .groups = "drop"
  )
View(cat_times)

Metadata_SBP2_baseline2 <- Metadata_SBP2_baseline %>%
  left_join(cat_times, by = "Cat_ID")

View(Metadata_SBP2_baseline2)

median(Metadata_SBP2_baseline2$diff_years)


quantile(Metadata_SBP2_baseline2$diff_years, 0.25)

quantile(Metadata_SBP2_baseline2$diff_years, 0.75)

min(Metadata_SBP2_baseline2$diff_years)

max(Metadata_SBP2_baseline2$diff_years)

```

##b. Is the SBP data normal?

```{r SBP normality}

hist(Metadata_SBP2$SBP, breaks=50)

qqnorm(Metadata_SBP2$SBP)
qqline(Metadata_SBP2$SBP)

shapiro.test(Metadata_SBP2$SBP)


##data is not normal 
```

##c. describe age at baseline (e.g. first SBP measured)
###i. create df of baseline data
```{r baseline df}
#create dataset of baseline data 

Metadata_SBP2_baseline<-Metadata_SBP2 %>%
  arrange(Cat_ID, date) %>%
  group_by(Cat_ID) %>%
  slice(1) %>%
  ungroup()

```
###ii. check normality
```{r baseline sbp normality}
#is it normal? 
hist(Metadata_SBP2_baseline$SBP, breaks=50)

qqnorm(Metadata_SBP2_baseline$SBP)
qqline(Metadata_SBP2_baseline$SBP)

shapiro.test(Metadata_SBP2_baseline$SBP)
#not normal
```

###iii.summarise baseline data
```{r}

#summarise age
summarise_age(Metadata_SBP2_baseline)

#average years follow up per cat
# Calculate follow-up duration per cat
followup_per_cat <- Metadata_SBP2 %>%
  group_by(Cat_ID) %>%
  summarise(
    followup_years = max(Age_years, na.rm = TRUE) - min(Age_years, na.rm = TRUE)
  )

print(followup_per_cat)

hist(followup_per_cat$followup_years, breaks=10)
shapiro.test(followup_per_cat$followup_years)

#not normal - median IQR range
# Calculate average follow-up duration across all cats
median_followup <- median(followup_per_cat$followup_years, na.rm = TRUE)
Q1_followup<- quantile(followup_per_cat$followup_years, 0.25)
Q3_followup<- quantile(followup_per_cat$followup_years, 0.75)
min_followup<- min(followup_per_cat$followup_years)
max_followup<- max(followup_per_cat$followup_years)

# Print results
print(followup_per_cat)       # follow-up per cat
cat("Median follow-up time (years):", median_followup, "\n")
cat("Q1 follow-up time (years):", Q1_followup, "\n")
cat("Q3 follow-up time (years):", Q3_followup, "\n")
cat("min follow-up time (years):", min_followup, "\n")
cat("max follow-up time (years):", max_followup, "\n")
```
###iv. describe clinical factors
```{r describe data}


median(Metadata_SBP2_baseline$SBP, na.rm=T)
range(Metadata_SBP2_baseline$SBP, na.rm=T)
IQR(Metadata_SBP2_baseline$SBP, na.rm=T)
quantile(Metadata_SBP2_baseline$SBP, 0.25, na.rm=T)
quantile(Metadata_SBP2_baseline$SBP, 0.75, na.rm=T)


median(Metadata_SBP2_baseline$weight, na.rm=T)
range(Metadata_SBP2_baseline$weight, na.rm=T)
IQR(Metadata_SBP2_baseline$weight, na.rm=T)
quantile(Metadata_SBP2_baseline$weight, 0.25, na.rm=T)
quantile(Metadata_SBP2_baseline$weight, 0.75, na.rm=T)

median(Metadata_SBP2_baseline$BCS, na.rm=T)
range(Metadata_SBP2_baseline$BCS, na.rm=T)
IQR(Metadata_SBP2_baseline$BCS, na.rm=T)
quantile(Metadata_SBP2_baseline$BCS, 0.25, na.rm=T)
quantile(Metadata_SBP2_baseline$BCS, 0.75, na.rm=T)

median(Metadata_SBP2_baseline$MCS_combined, na.rm=T)
range(Metadata_SBP2_baseline$MCS_combined, na.rm=T)
IQR(Metadata_SBP2_baseline$MCS_combined, na.rm=T)
quantile(Metadata_SBP2_baseline$MCS_combined, 0.25, na.rm=T)
quantile(Metadata_SBP2_baseline$MCS_combined, 0.75, na.rm=T)


#breed
sum(Metadata_SBP2_baseline$Breed=="Pedigree")
sum(Metadata_SBP2_baseline$Breed!="Pedigree")

Metadata_SBP2_baseline%>%
  group_by(breed)%>%
  summarise(N=n())
#sex
sum(Metadata_SBP2_baseline$Sex=="Male")
sum(Metadata_SBP2_baseline$Sex=="Female")
#neuter
sum(Metadata_SBP2_baseline$neuter_status=="No")

#morbidities at baseline
sum(Metadata_SBP2_baseline$CKD=="Yes")
sum(Metadata_SBP2_baseline$Hypertension=="Yes")
sum(Metadata_SBP2_baseline$diabetes_individual.tp=="Yes")
sum(Metadata_SBP2_baseline$Hyperthyroidism=="Yes")
sum(Metadata_SBP2_baseline$oe.abnormalities_Yes_no_incomplete=="Yes", na.rm=TRUE)
sum(Metadata_SBP2_baseline$Dental_disease=="Yes")
sum(Metadata_SBP2_baseline$Neoplasia=="Yes")
sum(Metadata_SBP2_baseline$FIV.Y.N=="Yes")
sum(Metadata_SBP2_baseline$Heart_murmur=="Yes")

#morbidities by study end
sum(Metadata_SBP2_baseline$CKD_over.study=="Yes")
sum(Metadata_SBP2_baseline$hypertensive_over.study=="Yes")
sum(Metadata_SBP2_baseline$Diabetes_mellitus=="Yes")
sum(Metadata_SBP2_baseline$hyperthyroid_over.study=="Yes")
sum(Metadata_SBP2_baseline$oe_abnormalities_over_study=="Yes")
sum(Metadata_SBP2_baseline$dental_dz_over_study=="Yes")
sum(Metadata_SBP2_baseline$neoplasia_over.study=="Yes")
sum(Metadata_SBP2_baseline$HM_at_least_once=="Yes")


```


###v. code for themed plots
```{r theme}
My_Theme = theme(
  axis.title.x = element_text(size = 16),
  axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size=14),
  axis.title.y = element_text(size = 16), 
 plot.title = element_text(size = 16, hjust = 0.5))
```

##d. plot changes in spahetti plot to see SBP
###i. ## age overall
```{r spaghetti plots}
g<-ggplot(Metadata_SBP2, aes(x = Age_years, y = SBP, group = Cat_ID)) +
  geom_line(aes(group=factor(Cat_ID), col=Cat_ID, alpha = 0.5)) +  # Add lines connecting data points for each cat
  geom_point(alpha = 0.5) +  # Add points for each data point
  labs(x = "Age (years)", y = "Systolic blood pressure (mmHg)", title = "Trends of SBP changes with age in 203 cats") +
  theme_minimal() +
  theme(legend.position="none")

g +My_Theme

#ggsave("Trends of SBP changes with age in 203 cats.png", #plot= last_plot(), bg = "white", height = 2100, width = #3150, units="px" )
```

###ii. sex
```{r spag plot sex}

g2<-ggplot(Metadata_SBP2, aes(x = Age_years, y = SBP, group = Cat_ID)) +
  geom_line(aes(group=factor(Cat_ID), col=Sex)) +  # Add lines connecting data points for each cat
  geom_point(alpha = 0.5) +  # Add points for each data point
  labs(x = "Age (years)", y = "Systolic blood pressure (mmHg)", title = "Trends of SBP changes with age and sex in 203 cats") +
  theme_minimal() +
  theme(legend.position="bottom")

g2 +My_Theme

#ggsave("Trends of SBP changes with age in 203 cats by #sex.png", plot= last_plot(), bg = "white", height = #2100, width = 3150, units="px" )

```

###iii. facet by Cat_ID
```{r}

# Split dataframe
cat_ids <- unique(Metadata_SBP2$Cat_ID)

first  <- cat_ids[1:56]
second <- cat_ids[57:112]
third <- cat_ids[113:168]
fourth<- cat_ids[169:203]

ggplot(filter(Metadata_SBP2, Cat_ID %in% first),
             aes(Age_years, SBP, group = Cat_ID)) +
  facet_wrap(~ Cat_ID, ncol = 7) +
  geom_line(alpha = 0.5, colour = "steelblue") +
  geom_point(alpha = 0.6, colour = "steelblue") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(filter(Metadata_SBP2, Cat_ID %in% second),
             aes(Age_years, SBP, group = Cat_ID)) +
  facet_wrap(~ Cat_ID, ncol = 7) +
  geom_line(alpha = 0.5, colour = "steelblue") +
  geom_point(alpha = 0.6, colour = "steelblue") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(filter(Metadata_SBP2, Cat_ID %in% third),
             aes(Age_years, SBP, group = Cat_ID)) +
  facet_wrap(~ Cat_ID, ncol = 7) +
  geom_line(alpha = 0.5, colour = "steelblue") +
  geom_point(alpha = 0.6, colour = "steelblue") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(filter(Metadata_SBP2, Cat_ID %in% fourth),
             aes(Age_years, SBP, group = Cat_ID)) +
  facet_wrap(~ Cat_ID, ncol = 7) +
  geom_line(alpha = 0.5, colour = "steelblue") +
  geom_point(alpha = 0.6, colour = "steelblue") +
  theme_minimal() +
  theme(legend.position = "none")
```

##e. Linear mixed effects models for SBP changes with age. 
###i. model 1 SBP, age, sex
####check splines
```{r check splines}
#fit model for no splines up to 5df and compare
mod_b1<-lmer(SBP ~ Age_years + Sex + Age_years:Sex +(1|Cat_ID), Metadata_SBP2)
mod_b2<-lmer(SBP ~ ns(Age_years, df=2) + Sex + ns(Age_years, df=2):Sex +(1|Cat_ID), Metadata_SBP2)
mod_b3<-lmer(SBP ~ ns(Age_years, df=3) + Sex + ns(Age_years, df=3):Sex +(1|Cat_ID), Metadata_SBP2)
mod_b4<-lmer(SBP ~ ns(Age_years, df=4) + Sex + ns(Age_years, df=4):Sex +(1|Cat_ID), Metadata_SBP2)
mod_b5<-lmer(SBP ~ ns(Age_years, df=5) + Sex + ns(Age_years, df=5):Sex +(1|Cat_ID), Metadata_SBP2)

#compare
anova(mod_b1, mod_b2, mod_b3, mod_b4, mod_b5)

#mod_b2 with 2df best

#check model assumptions

check_model(mod_b2)
acf(residuals(mod_b2))##ACF not showing stong autocorrelation. OK to use model. 
#png("check1.png", width = 2000, height = 3000, res = #300)
#print(check1)
#dev.off()
```

####add owner as nested re
```{r owner nested mod1}
mod_b2_nested<-lmer(SBP ~ ns(Age_years, df=2) + Sex + ns(Age_years, df=2):Sex +(1|Owner_ID/Cat_ID), Metadata_SBP2)
summary(mod_b2_nested)

tab_model(mod_b2_nested)

anova(mod_b2, mod_b2_nested)
```

####check mod1 assumptions
```{r check mod1}
check1<-check_model(mod_b2_nested)
print(check1)
png("check1.png", width = 2000, height = 3000, res = 300)
print(check1)
dev.off()

VarCorr(mod_b2_nested)
```

####compare overall effect of age with LRT
```{r LRT mode1}
mod_b2_noage<-lmer(SBP ~  Sex +  +(1|Owner_ID/Cat_ID), Metadata_SBP2)

anova(mod_b2_noage, mod_b2_nested)

#mod with age significantly improves model fit P<0.001
```

###ii. model2: sbp, age, sex, diseases
####ensure data ordered correctly
```{r order mcs}
Metadata_SBP2$MCS_grade_None_Mild_Moderate_to_Severe<-factor(Metadata_SBP2$MCS_grade_None_Mild_Moderate_to_Severe, levels= c( "None", "Mild", "Moderate to severe"))
```

#####arrange BCS into categories

```{r categorise bcs}
Metadata_SBP2 <- Metadata_SBP2 %>%
  mutate(BCS_group3 = case_when(
    BCS %in% 1:4 ~ "1-4",
    BCS == 5     ~ "5",
    BCS == 6     ~ "6",
    BCS %in% 7:9 ~ "7-9",
    TRUE ~ NA_character_
  ))
#set ref category as 5 
Metadata_SBP2  <- Metadata_SBP2  %>%
  mutate(BCS_group3 = factor(
    BCS_group3,
    levels = c("5", "1-4", "6", "7-9")
  ))
# Count cats in each category
cat_counts <- Metadata_SBP2  %>%
  count(BCS_group3)
cat_counts 

```
##### group HM
```{r group hm}
Metadata_SBP2<-Metadata_SBP2%>%
  mutate(
    HM_group = case_when(
      HM_grade == 0 ~ "None", 
      HM_grade %in% c(1,2) ~ "I-II/VI",
      HM_grade >=3 ~ "≥III/VI",
      TRUE ~ NA_character_
    ))
#    
levels(Metadata_SBP2$HM_group)
#
as.factor(Metadata_SBP2$HM_group)  
#
Metadata_SBP2$HM_group<-factor(Metadata_SBP2$HM_group, levels = c("None", "I-II/VI", "≥III/VI"))
```
#####group muscle loss
```{r binary mcs}
Metadata_SBP2<-Metadata_SBP2%>%
  mutate(
    MCS_binary= case_when(
       MCS_combined == 30 ~ "None", 
       MCS_combined %in% c(0:29) ~ "Muscle loss",
      TRUE ~ NA_character_
    ))
#    
levels(Metadata_SBP2$MCS_binary)
#
as.factor(Metadata_SBP2$MCS_binary)  
#
Metadata_SBP2$MCS_binary<-factor(Metadata_SBP2$MCS_binary, levels = c("None", "Muscle loss"))

# Count cats in each category
cat_counts <- Metadata_SBP2  %>%
  count(MCS_binary)
cat_counts 
```

####run model
```{r model 2}
mod_bc <- lmer(SBP ~ ns(Age_years, df = 2) + Sex + ns(Age_years, df=2):Sex + Breed + BCS_group3 + MCS_grade_None_Mild_Moderate_to_Severe +
                 CKD + Hyperthyroidism +  Neoplasia + 
                 Heart_murmur + Diabetes_mellitus + Dental_disease + oe.abnormalities_Yes_no_incomplete +
                 (1 | Cat_ID), data = Metadata_SBP2)

summary(mod_bc)
```
#####check N per disease in this model
```{n per disease mod 2}
Metadata_mod_bc<- Metadata_SBP2 %>%
  select(Cat_ID, Cat_ID_timepoint, Sex, BCS_group, MCS_grade_None_Mild_Moderate_to_Severe, SBP, CKD, Hyperthyroidism, 
         Neoplasia, Heart_murmur, Diabetes_mellitus, 
         Dental_disease, oe.abnormalities_Yes_no_incomplete)
#remove NAs
Metadata_mod_bc<-na.omit(Metadata_mod_bc)

distinct_cats_per_covariate_mod_bc <- Metadata_mod_bc %>%
  summarise(
    BCS_underweight=n_distinct(Cat_ID[BCS_group == "Underweight"]), 
     BCS_ideal=n_distinct(Cat_ID[BCS_group == "Ideal weight"]), 
    BCS_overweight=n_distinct(Cat_ID[BCS_group == "Overweight"]), 
    BCS_obese=n_distinct(Cat_ID[BCS_group == "Obese"]), 
    MCS_none=n_distinct(Cat_ID[MCS_grade_None_Mild_Moderate_to_Severe == "None"]), 
    MCS_mild=n_distinct(Cat_ID[MCS_grade_None_Mild_Moderate_to_Severe == "Mild"]), 
    MCS_mod_to_severe=n_distinct(Cat_ID[MCS_grade_None_Mild_Moderate_to_Severe == "Moderate_to_severe"]), 
    CKD = n_distinct(Cat_ID[CKD == "Yes"]),
    Hyperthyroidism = n_distinct(Cat_ID[Hyperthyroidism == "Yes"]),
    Neoplasia = n_distinct(Cat_ID[Neoplasia == "Yes"]),
    Heart_murmur = n_distinct(Cat_ID[Heart_murmur == "Yes"]),
    diabetes = n_distinct(Cat_ID[Diabetes_mellitus == "Yes"]),
    Dental_disease = n_distinct(Cat_ID[Dental_disease == "Yes"]),
    oe_abnormalities_Yes = n_distinct(Cat_ID[oe.abnormalities_Yes_no_incomplete =="Yes"]),
    oe_abnormalities_unable = n_distinct(Cat_ID[oe.abnormalities_Yes_no_incomplete =="Incomplete_OE"]),
     Sex = n_distinct(Cat_ID[Sex == "Male"])
  )

#Print the number of distinct cats that meet the conditions
print(distinct_cats_per_covariate_mod_bc)

tbl_regression(mod_bc, exp=TRUE)%>%
  add_n(location="level")

tbl_regression(mod_bc2, exp=TRUE)%>%
  add_n(location="level")

tbl_regression(mod_bb1, exp=TRUE)%>%
  add_n(location="level")
```

#####owner nested
```{r owner nested mod2}
mod_bc_nested <- lmer(SBP ~ ns(Age_years, df = 2) + Sex + ns(Age_years, df=2):Sex + Breed + BCS_group3 + MCS_grade_None_Mild_Moderate_to_Severe +
                 CKD + Hyperthyroidism +  Neoplasia +
                 HM_group + HR + diabetes_individual.tp + Dental_disease  + (1 | Owner_ID/Cat_ID), data = Metadata_SBP2)

tab_model(mod_bc_nested)

```

#####compare splines
```{r compare spline mod 2}
mod_bc_nested2 <- lmer(SBP ~ Age_years + Sex + Age_years:Sex + Breed + BCS_group3 + MCS_grade_None_Mild_Moderate_to_Severe +
                 CKD + Hyperthyroidism +  Neoplasia +
                 HM_group + HR + diabetes_individual.tp + Dental_disease  + (1 | Owner_ID/Cat_ID), data = Metadata_SBP2)


anova(mod_bc_nested, mod_bc_nested2)#splines significantly improve model 
AIC(mod_bc_nested, mod_bc_nested2)
```

#####check mod 2 assumptions
```{r check_model2}

check2<-check_model(mod_bc_nested2)
print(check2)
check_collinearity(mod_bc_nested2)

png("check2.png", width = 2000, height = 3000, res = 300)
print(check2)
dev.off()
```
#####LRT to compare overall effect of age

```{r LRT mod2}
mod_bc_noage<- lmer(SBP ~  + Sex + Breed + BCS_group3 + MCS_grade_None_Mild_Moderate_to_Severe +
                 CKD + Hyperthyroidism +  Neoplasia +
                 HM_group + HR + diabetes_individual.tp + Dental_disease  + (1 | Owner_ID/Cat_ID), data = Metadata_SBP2)

anova(mod_bc_noage, mod_bc_nested)
#Age significantly improved model P<0.001
```

##iii. mod 3 subset for OE score
```{r mod 3}
mod_oe_nested <- lmer(SBP ~ ns(Age_years, df = 2) + Sex + ns(Age_years, df=2):Sex + Breed + BCS_group3 + MCS_grade_None_Mild_Moderate_to_Severe +
                 CKD + Hyperthyroidism +  Neoplasia +
                 HM_group + HR + diabetes_individual.tp + Dental_disease + oe.abnormality_score +  (1 | Owner_ID/Cat_ID), data = Metadata_SBP2)

tab_model(mod_oe_nested)

```

#####check mod 3 assumptions
```{r check_model}

check3<-check_model(mod_oe_nested)
print(check3)
check_collinearity(mod_oe_nested)

png("check3.png", width = 2000, height = 3000, res = 300)
print(check3)
dev.off()

check_collinearity(mod_oe_nested)
```

##iv. sensitivity analyses
####Now subset by healthy cats only

```{r healthy visits}

Metadata_healthy<-Metadata_SBP2%>%
  filter(Healthy_visit=="Yes")

dim(Metadata_healthy)

Metadata_healthy%>%
  filter(Hypertension=="Yes")


mod_healthy<-lmer(SBP ~ ns(Age_years, df = 2) + Sex + ns(Age_years, df=2):Sex + (1|Owner_ID/Cat_ID), Metadata_healthy)
mod_healthy2<-lmer(SBP ~ ns(Age_years, df = 3) + Sex + ns(Age_years, df=3):Sex + (1|Owner_ID/Cat_ID), Metadata_healthy)
tab_model(mod_healthy2)
mod_healthy3<-lmer(SBP ~ Age_years + Sex + Age_years:Sex + (1|Owner_ID/Cat_ID), Metadata_healthy)

anova(mod_healthy, mod_healthy2, mod_healthy3)


tab_model(mod_healthy)
plot_model(mod_healthy, type="pred", terms=c("Age_years[all]", "Sex"), show.data=F)
plot_model(mod_healthy, type="pred", terms=c("Age_years[all]"), show.data=F)
```

#####check mod assumptions
```{r check_model4}

check4<-check_model(mod_healthy)
print(check4)
check_collinearity(mod_healthy)
```

####subet by "healthy over entire study" interactions

```{r healthy over study interactions}
data_healthyoverstudy<-Metadata_SBP2%>%
  filter(Overall_health_status=="Healthy")
  

mod_healthyoverstudy<-lmer(SBP ~ ns(Age_years, df=2) + Sex + ns(Age_years, df=2):Sex +(1|Cat_ID), data_healthyoverstudy)

tab_model(mod_healthyoverstudy)

plot_model(mod_healthyoverstudy, type="pred")

#interactions
mod_int<-lmer(SBP ~ ns(Age_years, df=2) + Sex + ns(Age_years, df=2):Sex + Overall_health_status + ns(Age_years, df=2):Overall_health_status+ (1|Owner_ID/Cat_ID), Metadata_SBP2)

tab_model(mod_int)

plot_model(mod_int, type="emm", terms=c("Age_years", "Overall_health_status"), colors = c("orange", "purple"))+
  xlab("Age (years)") +
  ylab("SBP (mmHg)") +
  ggtitle("") +
  theme_bw() +
  theme(legend.position = "none")+
  scale_y_continuous(limits = c(100, 200))

Metadata_SBP2%>% filter (Overall_health_status=="Healthy") %>%   summarise(n_healthy_cats = n_distinct(Cat_ID))
#
ggsave("Fig3_age_health_interactions.png", plot= last_plot(), bg = "white", height =10, width = 15, units="cm" )

```
#####Check mod assumptions
```{r check_model4}

check4<-check_model(mod_int)
png("check4.png", width = 3000, height = 4000, res = 300)
print(check4)
dev.off()

check_collinearity(mod_int)

```

##v. table of results

#####tables for nested models
```{r tab LMMs}
tab_model(mod_b2_nested, mod_bc_nested, mod_oe_nested, mod_healthy)

```

##vi. plot of results
```{r plots LMMs}

pA2<-plot_model(mod_b2_nested, type = "emm", color = "black", terms = c("Age_years[all]")) +
  xlab("Age (years)") +
  ylab("SBP (mmHg)") +
  ggtitle("A") +
  theme_bw() +
  scale_y_continuous(limits = c(100, 200))
pA2
#ggsave("A SBP and age plot_mod1.png", plot= last_plot(), bg = "white", height =10, width = 10, units="cm" )

pB2<-plot_model(mod_b2_nested, type="emm",color=c("#00BFC4", "#F8766D"), terms=c("Age_years[all]", "Sex"))+ 
  xlab("Age (years)")+
  ylab("SBP (mmHg)") + 
  ggtitle("B")+  
  theme_bw()+
  theme(legend.position = "none")+
  scale_y_continuous(limits = c(100, 200))

pB2
#ggsave("B SBP and sex and age plot_mod1.png", plot= last_plot(), bg = "white", height =10, width = 13, units="cm" )

pC2<-plot_model(mod_healthy, type = "emm", color = "black", terms = c("Age_years[all]")) +
  xlab("Age (years)") +
  ylab("SBP (mmHg)") +
  ggtitle("A") +
  theme_bw() +
  scale_y_continuous(limits = c(100, 200))

#ggsave("C SBP and age plot_healthy cats only_mod4.png", #plot= last_plot(), bg = "white", height =10, width = 10, #units="cm" )

pD2<-plot_model(mod_healthy, type="emm",color=c("#00BFC4", "#F8766D"), terms=c("Age_years[all]", "Sex"))+ 
  xlab("Age (years)")+
  ylab("SBP (mmHg)") + 
  ggtitle("B")+  
  theme_bw()+
  theme(legend.position="none")+
  scale_y_continuous(limits = c(100, 200))
#ggsave("D SBP and sex and age plot_healthy cats #only_mod4.png", plot= last_plot(), bg = "white", height #=10, width = 13, units="cm" )




#patchwork together 

plotsA2_B2<-(pA2 | pB2) 

ggsave("plotsA2_B2.png", plot=plotsA2_B2, bg = "white", height =10, width = 24, units="cm" )

plots_healthy2_bw<-(pC2 | pD2)


ggsave("plots_healthy2_bw.png", plot=plots_healthy2_bw,
       
plot_model(mod_int,
           type = "emm",
           color = c("orange", "purple"),
           terms = c("Age_years[all]", "Overall_health_status")) +
  xlab("Age (years)") +
  ylab("SBP (mmHg)") +
  ggtitle("") +
  theme_bw() +
  theme(
    legend.position = "none") +
  scale_y_continuous(limits = c(100, 200))
ggsave("Fig.3.SBP and sex and health plot.png", plot= last_plot(), bg = "white", height =10, width = 13, units="cm" )


```

##vii. emmeans
```{r}
library(emmeans)
em <- emmeans(mod_b2_nested, ~ Age_years,
              at = list(Age_years = c(6,7,8,9, 10, 10.5, 11, 12, 13, 14, 15, 16)),
              re.form = NA)  # Omit random effects (population-level prediction)

summary(em)

em_sex <- emmeans(mod_b2_nested, ~ Age_years + Sex,
              at = list(Age_years = c(6,7,8,9, 10, 10.5, 11, 12, 13, 14, 15, 16, 17)),
              re.form = NA)  # Omit random effects (population-level prediction)

summary(em_sex)

em_healthy <- emmeans(mod_healthy, ~ Age_years,
              at = list(Age_years = c(6,7,8,9, 10, 10.5, 11, 12, 13, 14, 15, 16, 17)),
              re.form = NA)  # Omit random effects (population-level prediction)

summary(em_healthy )

em_healthy_sex <- emmeans(mod_healthy, ~ Age_years + Sex,
              at = list(Age_years = c(6,7,8,9, 10, 10.5, 11, 12, 13, 14, 15, 16, 17)),
              re.form = NA)  # Omit random effects (population-level prediction)

summary(em_healthy_sex)
```

##viii. sample sizes per model
```{r}
#for visit numbers
> tbl_regression(mod_bc_nested, exp=T)%>%add_n(location="level")
> tbl_regression(mod_oe_nested, exp=T)%>%add_n(location="level")
tbl_regression(mod_healthy, exp=T)%>%add_n(location="level")
#for cat numbers?
#check N per disease in this model
Metadata_mod_bc_nested<- Metadata_SBP2 %>%
  select(Cat_ID, Cat_ID_timepoint, Sex, BCS_group3, CKD, Hyperthyroidism, 
         Hypertension, Neoplasia, HM_group, diabetes_individual.tp, 
         Dental_disease, HR, Breed, MCS_grade_None_Mild_Moderate_to_Severe)
#remove NAs
Metadata_mod_bc_nested<-na.omit(Metadata_mod_bc_nested)

distinct_cats_per_covariate_mod_bc_nested <- Metadata_mod_bc_nested %>%
  summarise(
    BCS5=n_distinct(Cat_ID[BCS_group3== "5"]),
     BCS14=n_distinct(Cat_ID[BCS_group3== "1-4"]),
     BCS6=n_distinct(Cat_ID[BCS_group3== "6"]),
     BCS79=n_distinct(Cat_ID[BCS_group3== "7-9"]),
     MCS3=n_distinct(Cat_ID[MCS_grade_None_Mild_Moderate_to_Severe== "None"]),
    MCS2=n_distinct(Cat_ID[MCS_grade_None_Mild_Moderate_to_Severe== "Mild"]),
    MCS01=n_distinct(Cat_ID[MCS_grade_None_Mild_Moderate_to_Severe== "Moderate to severe"]),
    CKD = n_distinct(Cat_ID[CKD == "Yes"]),
    Hyperthyroidism = n_distinct(Cat_ID[Hyperthyroidism == "Yes"]),
    Hypertension = n_distinct(Cat_ID[Hypertension == "Yes"]),
    Neoplasia = n_distinct(Cat_ID[Neoplasia == "Yes"]),
    HM_group0 = n_distinct(Cat_ID[HM_group == "None"]),
    HM_group12 = n_distinct(Cat_ID[HM_group  == "I-II/VI"]),
    HM_group3 = n_distinct(Cat_ID[HM_group  == "≥III/VI"]),
    diabetes_individual = n_distinct(Cat_ID[diabetes_individual.tp == "Yes"]),
    Dental_disease = n_distinct(Cat_ID[Dental_disease == "Yes"]),
     Sex = n_distinct(Cat_ID[Sex == "Male"]),
    Breed_ped=n_distinct(Cat_ID[Breed =="Pedigree"])
  )

#Print the number of distinct cats that meet the conditions
print(distinct_cats_per_covariate_mod_bc_nested)


```
####and for mod_oe
```{r}
#for cat numbers?
#check N per disease in this model
Metadata_mod_oe_nested<- Metadata_SBP2 %>%
  select(Cat_ID, Cat_ID_timepoint, Sex, BCS_group3, CKD, Hyperthyroidism, 
         Hypertension, Neoplasia, HM_group, diabetes_individual.tp, 
         Dental_disease, HR, Breed, MCS_grade_None_Mild_Moderate_to_Severe, oe.abnormality_score)
#remove NAs
Metadata_mod_oe_nested<-na.omit(Metadata_mod_oe_nested)

distinct_cats_per_covariate_mod_oe_nested <- Metadata_mod_oe_nested %>%
  summarise(
    BCS5=n_distinct(Cat_ID[BCS_group3== "5"]),
     BCS14=n_distinct(Cat_ID[BCS_group3== "1-4"]),
     BCS6=n_distinct(Cat_ID[BCS_group3== "6"]),
     BCS79=n_distinct(Cat_ID[BCS_group3== "7-9"]),
     MCS3=n_distinct(Cat_ID[MCS_grade_None_Mild_Moderate_to_Severe== "None"]),
    MCS2=n_distinct(Cat_ID[MCS_grade_None_Mild_Moderate_to_Severe== "Mild"]),
    MCS01=n_distinct(Cat_ID[MCS_grade_None_Mild_Moderate_to_Severe== "Moderate to severe"]),
    CKD = n_distinct(Cat_ID[CKD == "Yes"]),
    Hyperthyroidism = n_distinct(Cat_ID[Hyperthyroidism == "Yes"]),
    Hypertension = n_distinct(Cat_ID[Hypertension == "Yes"]),
    Neoplasia = n_distinct(Cat_ID[Neoplasia == "Yes"]),
    HM_group0 = n_distinct(Cat_ID[HM_group == "None"]),
    HM_group12 = n_distinct(Cat_ID[HM_group  == "I-II/VI"]),
    HM_group3 = n_distinct(Cat_ID[HM_group  == "≥III/VI"]),
    diabetes_individual = n_distinct(Cat_ID[diabetes_individual.tp == "Yes"]),
    Dental_disease = n_distinct(Cat_ID[Dental_disease == "Yes"]),
     Sex = n_distinct(Cat_ID[Sex == "Male"]),
    Breed_ped=n_distinct(Cat_ID[Breed =="Pedigree"])
  )

#Print the number of distinct cats that meet the conditions
print(distinct_cats_per_covariate_mod_oe_nested)


```
#####and for healthy

```{r}
#for cat numbers?
#check N per disease in this model
Metadata_mod_healthy<- Metadata_SBP2 %>%
  filter(Healthy_visit=="Yes")%>%
  select(Cat_ID, Cat_ID_timepoint, Sex, Age_years)
#remove NAs
Metadata_mod_healthy<-na.omit(Metadata_mod_healthy)

distinct_cats_per_covariate_mod_healthy<- Metadata_mod_healthy %>%
  summarise(
     SexM = n_distinct(Cat_ID[Sex == "Male"]),
     SexF = n_distinct(Cat_ID[Sex == "Female"]),
  )

#Print the number of distinct cats that meet the conditions
print(distinct_cats_per_covariate_mod_healthy)

dim(Metadata_mod_healthy)
```

#7. model visits where SBP >=160mmHg
##a. binomial model of hypertension
```{r}
Metadata_SBP2$Hypertension<-as.factor(Metadata_SBP2$Hypertension.y)

Metadata_209_cats$Hypertension<-as.factor(Metadata_209_cats$Hypertension)
#full model

mod_ht<-glmer(Hypertension ~ Age_years + Sex + Breed + BCS_group2 + MCS_grade_None_Mild_Moderate_to_Severe + CKD + Hyperthyroidism + Neoplasia + diabetes_individual.tp + HM_group + Dental_disease + (1|Cat_ID), data= Metadata_SBP2, family=binomial)

#
summary(mod_ht)
tab_model(mod_ht)
tbl_regression(mod_ht, exp=T)%>% add_n(location="level")
####model fails to converge? neoplasia infinite

#remove neoplasia and diabetic cases as too few samples
Metadata_SBP_nodmorneo<- Metadata_SBP2 %>%
  filter(diabetes_individual.tp != "Yes")

Metadata_SBP_nodmorneo<- Metadata_SBP_nodmorneo %>%
  filter(Neoplasia != "Yes")

#re-run model

mod_ht<-glmer(Hypertension ~ Age_years + Sex + Breed + HR + BCS_group2 + MCS_grade_None_Mild_Moderate_to_Severe + CKD + Hyperthyroidism + HM_group + Dental_disease + (1|Cat_ID), data= Metadata_SBP_nodmorneo, family=binomial) #failed to converge

mod_ht<-glmer(Hypertension ~ Age_years + Sex + BCS_group2 + MCS_grade_None_Mild_Moderate_to_Severe + CKD + Hyperthyroidism + Heart_murmur+ Dental_disease + (1|Cat_ID), data= Metadata_SBP_nodmorneo, family=binomial) #failed to converge

#
summary(mod_ht)
tab_model(mod_ht)
tbl_regression(mod_ht, exp=T)%>% add_n(location="level")

mod_ht<-glmer(Hypertension ~ Age_years + Sex + Breed + HR + BCS_group2 + MCS_grade_None_Mild_Moderate_to_Severe + CKD + Hyperthyroidism + HM_group + Dental_disease + (1|Cat_ID), data= Metadata_SBP_nodmorneo, family=binomial) #failed to converge

#
summary(mod_ht)
tab_model(mod_ht)
tbl_regression(mod_ht, exp=T)%>% add_n(location="level")

```


##b. GEE models for SBP >160 appointments
###i. sort data
```{r data_for_gee}
#create dataframe without NAs

Metadata_SBP_gee<-na.omit(Metadata_SBP2[ ,c("Hypertension_num", "Cat_ID", "Owner_ID", "Age_years", "Sex", "Breed", "BCS_group3", "MCS_binary", "HR", "CKD", "diabetes_individual.tp", "Neoplasia", "SBP", "Hyperthyroidism", "Heart_murmur", "HR", "Dental_disease")])

sum(Metadata_SBP_gee$Hypertension_num)

Metadata_SBP_gee$hypertensive_over.study_num<-Metadata_SBP_gee$hypertensive_over.study
sum(Metadata_SBP_gee$hypertensive_over.study_num)

length(unique(Metadata_SBP_gee$Cat_ID[Metadata_SBP_gee$Hypertension_num=="1"]))

#remove dm and neo

#Metadata_SBP_gee<-na.omit(Metadata_SBP_nodmorneo[ ,c("Hypertension_num", "Cat_ID", "Owner_ID", "Age_years", "Sex", "Breed", "BCS_group2", #"MCS_grade_None_Mild_Moderate_to_Severe", "HR", "CKD", "SBP", "Hyperthyroidism","HM_group", "Heart_murmur", "HR", "Dental_disease", "Healthy_visit")])
```

###ii. try GEE package
```{r gee package mod}


gee_ht<-gee(Hypertension_num ~ Sex + Breed, 
            data=Metadata_SBP2,
            id= Cat_ID,
            family=binomial, 
            corstr="exhangeable")
#won't work as some cats only have 1 measurement. Try geeglm
```

###iii. try geepack glm model
```{r geepack}


gee_ht <- geeglm(
  Hypertension_num ~ Age_years + Sex + Breed + HR + BCS_group3 + MCS_binary + Neoplasia + diabetes_individual.tp +
    CKD + Hyperthyroidism +  Heart_murmur + Dental_disease,
  data = Metadata_SBP_gee,
  id = Cat_ID,
  family = binomial,
  corstr = "exchangeable"
)

tbl_regression(gee_ht, exp=TRUE)%>%add_n(location="level")

tab_model(gee_ht)

summary(gee_ht)
```
####check mod assumptions
```{r check_geeglm}

check_collinearity(gee_ht)
check_residuals(gee_ht)


```

####check neoplasia results
```{r check_neo}
with(Metadata_SBP_gee, table(Neoplasia, Hypertension_num))

with(Metadata_SBP_gee,
     table(Neoplasia, tapply(Hypertension_num, Cat_ID, max)))


length(unique(Metadata_SBP_gee$Cat_ID[Metadata_SBP_gee$Neoplasia=="Yes"]))

#Zero hypertensive outcomes among neoplasia-positive cats
```


####remove neoplasia Neoplasia was excluded from multivariable models due to quasi-complete separation (no hypertensive outcomes observed among exposed cats), resulting in unstable estimates.
```{r no neo}
gee_ht2 <- geeglm(
  Hypertension_num ~ Age_years + Sex + Breed + HR + BCS_group3 + MCS_binary + diabetes_individual.tp +
    CKD + Hyperthyroidism +  Heart_murmur + Dental_disease,
  data = Metadata_SBP_gee,
  id = Cat_ID,
  family = binomial,
  corstr = "exchangeable"
)

tbl_regression(gee_ht2, pvalue_fun = label_style_pvalue(digits = 3), exp=TRUE)%>%add_n(location="level")
summary(gee_ht2)

```


####check over-parameterised
```{r check_geeglm}
table(Metadata_SBP_gee$Hypertension_num)
# Count of events
length(coef(gee_ht2))
# Number of estimated parameters
#EPV=7

summary(gee_ht2)$coefficients[, "Std.err"]

#correlation

vif(lm(Hypertension_num ~ Age_years + Sex + Breed + HR + BCS_group3 + MCS_binary + diabetes_individual.tp +
    CKD + Hyperthyroidism +  Heart_murmur + Dental_disease, data = Metadata_SBP_gee))

check_model(gee_ht2)
check_collinearity(gee_ht2)
check_homogeneity(gee_ht2)
```

##iii. GEE2 for OE score also

```{r geeglm oe}
Metadata_SBP_gee_oe<-na.omit(Metadata_SBP2[ ,c("Cat_ID", "Owner_ID", "Age_years", "Sex", "Breed", "BCS_group3", "MCS_binary", "HR", "CKD", "Hypertension_num", "SBP", "diabetes_individual.tp", "Hyperthyroidism", "Heart_murmur", "HR", "Dental_disease", "oe.abnormality_score", "Healthy_visit")])

sum(Metadata_SBP_gee_oe$Hypertension_num)
length(unique(Metadata_SBP_gee_oe$Cat_ID[Metadata_SBP_gee_oe$Hypertension_num=="1"]))

gee_ht_oe <- geeglm(
  Hypertension_num ~ Age_years + Sex + Breed + HR + BCS_group3 + MCS_binary + 
    CKD + Hyperthyroidism  + Heart_murmur + diabetes_individual.tp  + Dental_disease +oe.abnormality_score,
  data = Metadata_SBP_gee_oe,
  id = Cat_ID,
  family = binomial,
  corstr = "exchangeable"
)

#tab model

tbl_regression(gee_ht_oe,  pvalue_fun = label_style_pvalue(digits = 3), exp=TRUE)%>%add_n(location="level")
tab_model(gee_ht_oe)

#check model
check_collinearity(gee_ht_oe)
check_model(gee_ht_oe)

```

#8. TTE - Cox PH model

##a. sort data
###i. create baseline dataset

```{r baseline_data}

Metadata_hyp_baseline <- Metadata_SBP2 %>%
  arrange(Cat_ID, date) %>%
  group_by(Cat_ID) %>%
  slice(1) %>%
  ungroup()
```

###ii. Merge new incidence_data looking at censoring at age at last study visit NOT last study date
```{r read incidence data}

Hypertension_incidence_data2170326 <- read.csv("Hypertension_incidence_data2170326.csv")

Metadata_hyp_baseline2 <- Hypertension_incidence_data2170326 %>%
  left_join(
    Metadata_SBP2 %>%
      select(
        Cat_ID_timepoint,
        BCS_group2,
        BCS_group3,
        Heart_murmur,
        HR,
        CKD,
        Hyperthyroidism,
        Dental_disease,
        oe.abnormalities_Yes_no_incomplete,
        MCS_grade_None_Mild_Moderate_to_Severe
      ),
    by = "Cat_ID_timepoint"
  )

nrow(Metadata_hyp_baseline2)
colnames(Metadata_hyp_baseline2)
Metadata_hyp_baseline2<- Metadata_hyp_baseline2 %>%
  rename(SBP = hypertensive_sbp)

#no BCS 1-2 in baseline - reword categories
Metadata_hyp_baseline2 <- Metadata_hyp_baseline2 %>%
  mutate(BCS_group3 = case_when(
    BCS %in% 3:4 ~ "3-4",
    BCS == 5     ~ "5",
    BCS == 6     ~ "6",
    BCS %in% 7:9 ~ "7-9",
    TRUE ~ NA_character_
  ))
#set ref category as 5 
Metadata_hyp_baseline2  <- Metadata_hyp_baseline2 %>%
  mutate(BCS_group3 = factor(
    BCS_group3,
    levels = c("5", "3-4", "6", "7-9")
  ))

```

####iii. remove those cats diagnosed at baseline
```{r remove_baseline_dx}
Metadata_hyp_baseline2<-Metadata_hyp_baseline2%>%
  filter(baseline_dx =="No")
```

##b. Check MCS missingness
```{r MCS}

#there are 11 missing MCS at baseline. does MCS affect hypertension risk?

cox_mcs<-coxph(Surv(Censor_visit_age, stHyp ) ~ MCS_grade_None_Mild_Moderate_to_Severe,
                    data = Metadata_hyp_baseline2)

summary(cox_mcs)#not significant

```

##c. create complete case dataset

```{r complete_data}
#complete cases

Metadata_hyp_baseline_for_cox<-na.omit(Metadata_hyp_baseline2[,c("Cat_ID", "Age_years", "Censor_visit_age", "stHyp", "Sex", "SBP", "Breed", "BCS_group2", "BCS_group3")])

dim(Metadata_hyp_baseline_for_cox)

median(Metadata_hyp_baseline_for_cox$Age_years)
quantile(Metadata_hyp_baseline_for_cox$Age_years, 0.25)
quantile(Metadata_hyp_baseline_for_cox$Age_years, 0.75)
min(Metadata_hyp_baseline_for_cox$Age_years)
max(Metadata_hyp_baseline_for_cox$Age_years)

Metadata_hyp_baseline_for_cox%>%
  filter(stHyp==2)
```

##d. examine healthy visits only

```{r healthy}
#exmaine healthy visits only?

#merge healthy_visit
Metadata_hyp_baseline_3 <- merge(
  Metadata_hyp_baseline2,
  Metadata_SBP2[, c("Cat_ID_timepoint", "Healthy_visit")],
  by = "Cat_ID_timepoint",
  all = FALSE
)
#make healthy the reference
Metadata_hyp_baseline_3  <- Metadata_hyp_baseline_3 %>%
  mutate(Healthy_visit= factor(
   Healthy_visit,
    levels = c("Yes", "No" )
  ))

colnames(Metadata_hyp_baseline_3)
```
##e. inital Cox model exploring adding morbidities
```{r cox_morbs}
cox_1<-coxph(Surv(Censor_visit_age, stHyp ) ~ Sex + Breed + SBP + BCS_group3 + Heart_murmur + HR + CKD + Hyperthyroidism, data =  Metadata_hyp_baseline2)

tbl_regression(cox_1, exp=T)%>% add_n(location="level")

# unable to model with hyperthyroidism - unstable coefficient. Remove? CKD also limitis event numbers

cox_2<-coxph(Surv(Censor_visit_age, stHyp ) ~ Sex + Breed + BCS_group3 + MCS_grade_None_Mild_Moderate_to_Severe + Heart_murmur +CKD +Dental_disease + oe.abnormalities_Yes_no_incomplete, data = Metadata_hyp_baseline2)

tbl_regression(cox_2, exp=T)%>% add_n(location="level")

#strip down to healthy visit rather tahn co-morbs

cox_3<-coxph(Surv(Censor_visit_age, stHyp ) ~ Sex + Breed + BCS_group3 + MCS_grade_None_Mild_Moderate_to_Severe + Healthy_visit, data = Metadata_hyp_baseline_3)

tbl_regression(cox_3, exp=T)%>% add_n(location="level")

cox.zph(cox_3)

##issues - low event numbers of certain morbidities, healthy overall creates bias as unhealthy cats likely die earlier and are censored

```

##f. take healthy visit at baseline data only

```{r healthy_COX}
cox_healthy_data<-Metadata_hyp_baseline_3%>%
  filter(Healthy_visit=="Yes")
cox_healthy<-coxph(Surv(Censor_visit_age, stHyp ) ~ Sex + Breed + BCS_group3 + MCS_grade_None_Mild_Moderate_to_Severe, data = cox_healthy_data)

tbl_regression(cox_healthy, exp=T)%>% add_n(location="level")

#low sample size!
```

##g. forget about morbidities. Model using clinical data

```{r}
#all
cox_hyprisk<-coxph(Surv(Censor_visit_age, stHyp ) ~ Sex + Breed + SBP + 
                     BCS_group3,
                   data = Metadata_hyp_baseline_for_cox)

tbl_regression(cox_hyprisk, exp=T)%>%add_n(location=c("level"))
AIC(cox_hyprisk)


#remove breed
cox_hyprisk2<-coxph(Surv(Censor_visit_age, stHyp ) ~ Sex + SBP + 
                     BCS_group3,
                   data = Metadata_hyp_baseline_for_cox)

tbl_regression(cox_hyprisk2, exp=T)%>%add_n(location=c("level"))
AIC(cox_hyprisk2)


#SBP alone
cox_hyprisk3<-coxph(Surv(Censor_visit_age, stHyp ) ~ SBP,
                   data = Metadata_hyp_baseline_for_cox)

tbl_regression(cox_hyprisk3, exp=T)%>%add_n(location=c("level"))
AIC(cox_hyprisk3)



AIC(cox_hyprisk, cox_hyprisk2, cox_hyprisk3)#cox_2 best fit


summary(cox_hyprisk2)
```

###i. BEST MODEL
```{r best}
summary(cox_hyprisk2) 
tbl_regression(cox_hyprisk2, pvalue_fun= label_style_pvalue(digits = 3), exp=T)%>%add_n(location=c("level"))
```
###ii. check PH assumptions
```{r check_PH}
cox.zph(cox_hyprisk2)
plot(cox.zph(cox_hyprisk2))
```

###iii. check outliers
```{r check_outliers}
# Standardized dfbeta residuals
residuals(cox_hyprisk2, type = "dfbeta")
plot(residuals(cox_hyprisk2, type = "dfbeta"))  # look for points far from 0

# Deviance residuals
plot(residuals(cox_hyprisk2, type = "deviance"))

```
###iv. check linearity of SBP
```{r check_cox_splines}


# Fit model with a penalized spline for SBP
cox_spline <- coxph(Surv(Censor_visit_age, stHyp) ~ ns(SBP, df=2) + Sex + BCS_group3,
                    data = Metadata_hyp_baseline_for_cox)
summary(cox_spline)  # or compare log-likelihood to linear model
AIC(cox_hyprisk2, cox_spline)
#some slight imp to model fit but ideally keep the simpler linear model for easier interpretation
```

###v. assess concordance and c-index
```{r concordance}
#Concordance = 0.730
concordance(cox_hyprisk2)$concordance
summary(cox_hyprisk2)

```
###vi. check for multicollinearity

```{r vif_cox}

vif(cox_hyprisk2)
#all VIF<2
```

### vii. bootstrap 

```{r bootstrap}


# Function to compute coefficients in bootstrap sample
cox_boot_fun <- function(data, indices) {
  d <- data[indices, ]  # resample with replacement
  fit <- coxph(Surv(Censor_visit_age, stHyp) ~ Sex + SBP + BCS_group3, data = d)
  return(coef(fit))  # return coefficients
}

# Run bootstrap
set.seed(123)
boot_results <- boot(data = Metadata_hyp_baseline_for_cox,
                     statistic = cox_boot_fun,
                     R = 1000)  # number of bootstrap samples

# Inspect results
boot_results

# Bootstrap mean, SD, and 95% CI
apply(boot_results$t, 2, mean)      # mean coefficients
apply(boot_results$t, 2, sd)        # SD of coefficients

# 95% percentile CI
apply(boot_results$t, 2, function(x) quantile(x, probs = c(0.025, 0.975)))


exp(apply(boot_results$t, 2, mean))  # mean HRs
exp(apply(boot_results$t, 2, function(x) quantile(x, probs=c(0.025,0.975))))  # 95% CI
```
#### run rms
```{r}

# Setup for rms functions
dd <- datadist(Metadata_hyp_baseline_for_cox)
options(datadist="dd")

cox_rms <- cph(Surv(Censor_visit_age, stHyp) ~ Sex + SBP + BCS_group3,
               data = Metadata_hyp_baseline_for_cox, x=TRUE, y=TRUE, surv=TRUE)

# Bootstrap validation
set.seed(123)
validate(cox_rms, B=1000, method="boot")  # B=1000 bootstrap samples

```

##vii. plot km curve

```{r km}

#KM
km1<-survfit2(Surv(Censor_visit_age, stHyp ) ~ Sex, data = Metadata_hyp_baseline_for_cox)%>%
  ggsurvfit() +
  labs(x = "Age (years)", y="Hypertension-free survival probability")+
  ggtitle("A. Hypertension-free survival by sex")+
  #add_risktable()+
  add_confidence_interval()+
  add_censor_mark( size=1)+
  scale_x_continuous(breaks = 6:17, limits = c(6, 17))+
theme(
  legend.position = "inside",
  legend.position.inside = c(0.15, 0.15),
  legend.text=element_text(size=9),
  legend.box.background =  element_rect(colour="black", linewidth = 1),
  legend.title = element_text(size=9), 
  legend.title.position="top"
)
km1
#ggsave("Hypertension_sex_KM_curev170326.png", #plot=last_plot(),height = 2100, width = 2100, #units="px" )


#KM
km2<-survfit2(Surv(Censor_visit_age, stHyp ) ~ BCS_group3, data = Metadata_hyp_baseline_for_cox)%>%
  ggsurvfit() +
  labs(x = "Age (years)", y="Hypertension-free survival probability", color="BCS")+
  ggtitle("B. Hypertension-free survival by BCS")+
 # add_risktable()+
  #add_confidence_interval()+
  scale_color_manual(values=c("3-4" = "black", "5" = "brown", "6" = "aquamarine4", "7-9" = "grey"))+ 
  add_censor_mark( size=1)+
  scale_x_continuous(breaks = 6:17, limits = c(6, 17))+
  theme(
  legend.position = "inside",
  legend.position.inside = c(0.1, 0.2),
  legend.text=element_text(size=9),
  legend.box.background =  element_rect(colour="black", linewidth = 1),
  legend.title = element_text(size=9)
)
km2
#ggsave("Hypertension_BCS_KM_curev170326.png", #plot=last_plot(),height = 2100, width = 2100, #units="px" )

km1+km2
#ggsave("Hypertension_sex_bcs_KM_curve_Figure170326.png"#, plot=last_plot(),height = 15, width = 30, units="cm" #)

```

###viii. plot cumulativehaz

```{r cumhaz}
#
ci1<-survfit2(Surv(end_age_in_years, stHyp ) ~ Sex, data = Metadata_hyp_baseline_for_cox)%>%
  ggsurvfit(type="cumhaz") +
  labs(x = "Age (years)")+
  ggtitle("A.Hypertension Diagnosis by Sex")+
  #add_risktable()+
  add_confidence_interval()+
  add_censor_mark( size=1)+
  scale_x_continuous(breaks = 6:17, limits = c(6, 17))+
theme(
  legend.position = "inside",
  legend.position.inside = c(0.1, 0.9),
  legend.text=element_text(size=9),
  legend.box.background =  element_rect(colour="black", linewidth = 1),
  legend.title = element_text(size=9), 
  legend.title.position="top"
)
ci1

#ggsave("Hypertension_sex_cumhaz21012025.png", #plot=last_plot(),height = 2100, width = 2100, #units="px" )


#
ci2<-survfit2(Surv(end_age_in_years, stHyp ) ~ BCS_group3, data = Metadata_hyp_baseline_for_cox)%>%
  ggsurvfit(type="cumhaz") +
  labs(x = "Age (years)", color="BCS")+
  ggtitle("B.Hypertension Diagnosis by BCS")+
 # add_risktable()+
  #add_confidence_interval()+
  scale_color_manual(values=c("3-4" ="black", "5" = "orange", "6" = "purple", "7-9" = "turquoise"))+ 
  add_censor_mark( size=1)+
  scale_x_continuous(breaks = 6:17, limits = c(6, 17))+
  theme(
  legend.position = "inside",
  legend.position.inside = c(0.1, 0.8),
  legend.text=element_text(size=9),
  legend.box.background =  element_rect(colour="black", linewidth = 1),
  legend.title = element_text(size=9)
)
ci2
#ggsave("Hypertension_BCS_cumhazv210126.png", #plot=last_plot(),height = 2100, width = 2100, #units="px" )

ci1+ci2
#ggsave("Hypertension_sex_bcs_cumhaz_Figure210126.png", #plot=last_plot(),height = 15, width = 30, units="cm" )
```

###ix. plot cumulative incidence

```{r cuminc}
#
ci1<-survfit2(Surv(end_age_in_years, stHyp ) ~ Sex, data = Metadata_hyp_baseline)%>%
  ggsurvfit(type="risk") +
  labs(x = "Age (years)")+
  ggtitle("A.Hypertension Diagnosis by Sex")+
  #add_risktable()+
  add_confidence_interval()+
  add_censor_mark( size=1)+
  scale_x_continuous(breaks = 6:17, limits = c(6, 17))+
theme(
  legend.position = "inside",
  legend.position.inside = c(0.1, 0.9),
  legend.text=element_text(size=9),
  legend.box.background =  element_rect(colour="black", linewidth = 1),
  legend.title = element_text(size=9), 
  legend.title.position="top"
)

#ggsave("Hypertension_sex_cuminc04122025.png", #plot=last_plot(),height = 2100, width = 2100, #units="px" )


#
ci2<-survfit2(Surv(end_age_in_years, stHyp ) ~ BCS_group2, data = Metadata_hyp_baseline)%>%
  ggsurvfit(type="risk") +
  labs(x = "Age (years)", color="BCS")+
  ggtitle("B.Hypertension Diagnosis by BCS")+
 # add_risktable()+
  #add_confidence_interval()+
  #scale_color_manual(values=c("Obese" = "orange", "Overweight" = "purple", "Ideal weight" = "turquoise"))+ 
  add_censor_mark( size=1)+
  scale_x_continuous(breaks = 6:17, limits = c(6, 17))+
  theme(
  legend.position = "inside",
  legend.position.inside = c(0.1, 0.8),
  legend.text=element_text(size=9),
  legend.box.background =  element_rect(colour="black", linewidth = 1),
  legend.title = element_text(size=9)
)
#ggsave("Hypertension_BCS_cuminc041225.png", #plot=last_plot(),height = 2100, width = 2100, #units="px" )

ci1+ci2
#ggsave("Hypertension_cum_incidence_Figure.png", #plot=last_plot(),height = 15, width = 30, units="cm" )
```